---
title: "The Impact of Air Pollution on Public Health"
student_number: "C21412282"
author: "Prabhjeet Bissora"
programme_code: "TU856"
format: html
editor_options:
  chunk_output_type: console
  
output:
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: fill
runtime: shiny
---

# Introduction

Air pollution is a critical global issue affecting millions of lives. I will aim to investigate the relationship between air pollution levels and public health outcomes using multiple datasets. I aim to understand whether increased pollution correlates with lower life expectancies.

```{r, echo=FALSE}

# Load Libraries


library(tidyverse)
library(ggplot2)
library(dplyr)
library(plotly)
library(shiny)
library(readr)
library(DT)
library(httr)
library(jsonlite)
library(here)
library(flexdashboard)
library(scales)
library(patchwork)
library(viridis)


```

```{r, echo=FALSE}

mydata <- file.path("C:","Users", "prabhser", "OneDrive - Technological University Dublin", "Desktop", "Computer Science", "year 4", "sem 2", "visualising data", "assignment")

# load dataframes from respective 
air_df <- file.path(mydata, "air_pollution.csv")
health_df <- file.path(mydata, "health_outcomes.csv")

# load data from dataframes
air_pollution <- read_csv(air_df)
health_outcomes <- read_csv(health_df)

# Data cleaning and preparation
combined_data <- air_pollution %>%
  mutate(Country = trimws(tolower(Country))) %>%
  full_join(
    health_outcomes %>% 
      mutate(Country = trimws(tolower(Country))),
    by = "Country"
  ) %>%
  drop_na() %>%
  mutate(
    Region = factor(Region),
    Date = as.Date(Date),
    Pollution_Level = case_when(
      `AQI Value` < 50 ~ "Good",
      `AQI Value` < 100 ~ "Moderate",
      `AQI Value` < 150 ~ "Unhealthy for Sensitive Groups",
      `AQI Value` < 200 ~ "Unhealthy",
      TRUE ~ "Very Unhealthy"
    ) %>% factor(levels = c("Good", "Moderate", "Unhealthy for Sensitive Groups", "Unhealthy", "Very Unhealthy"))
  )

glimpse(combined_data)

```

```{r, echo=FALSE}

#| fig-height: 8
#| fig-width: 10
chart1 <- ggplot(combined_data, aes(x = `AQI Value`, y = LifeExpectancy, 
                                   color = Region, size = Population,
                                   text = paste("Country:", Country,
                                                "<br>Life Expectancy:", LifeExpectancy,
                                                "<br>AQI:", `AQI Value`))) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "black", linetype = "dashed") +
  scale_color_viridis_d() +
  scale_size_continuous(labels = comma) +
  labs(
    title = "Air Quality vs. Life Expectancy by Region",
    subtitle = "Higher pollution levels correlate with lower life expectancy",
    x = "Air Quality Index (Higher = Worse)",
    y = "Life Expectancy (Years)",
    color = "Region",
    size = "Population"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom",
        plot.title = element_text(face = "bold"))

ggplotly(chart1, tooltip = "text") %>%
  layout(hoverlabel = list(bgcolor = "white"),
         annotations = list(
           x = 0.5, y = -0.2,
           text = "Data shows a clear negative correlation between air quality and life expectancy",
           showarrow = FALSE,
           xref = "paper",
           yref = "paper"
         ))
```

\

```{r, echo=FALSE}
#| fig-height: 8
#| fig-width: 10
chart2 <- combined_data %>%
  ggplot(aes(x = Pollution_Level, y = FertilityRate, fill = Region,
             text = paste("Country:", Country,
                          "<br>Respiratory Cases:", FertilityRate))) +
  geom_boxplot(alpha = 0.8) +
  scale_fill_viridis_d() +
  labs(
    title = "Air Pollution Levels and Respiratory Diseases",
    subtitle = "Higher pollution categories show increased respiratory disease rates",
    x = "Air Pollution Level Category",
    y = "Respiratory Disease Cases per 100,000",
    fill = "Region"
  ) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(face = "bold"))

ggplotly(chart2, tooltip = "text") %>%
  layout(hoverlabel = list(bgcolor = "white"))
```

```{r}
#| fig-height: 8
#| fig-width: 12
chart3 <- combined_data %>%
  # Ensure Year is extracted correctly
  mutate(Year = lubridate::year(Date)) %>%
  # Group and summarize
  group_by(Year, Region) %>%
  summarise(
    Avg_AQI = mean(`AQI Value`, na.rm = TRUE),
    Avg_LifeExp = mean(LifeExpectancy, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  # Plot
  ggplot(aes(x = Year)) +
  geom_line(aes(y = Avg_AQI, color = "AQI"), size = 1) +
  geom_line(aes(y = Avg_LifeExp * 5, color = "Life Expectancy"), size = 1) +  # Scaled for visibility
  scale_y_continuous(
    name = "Average AQI",
    sec.axis = sec_axis(~ . / 5, name = "Average Life Expectancy")  # Adjusted scaling
  ) +
  scale_color_manual(values = c("AQI" = "red", "Life Expectancy" = "blue")) +
  labs(
    title = "Air Quality and Life Expectancy Over Time",
    x = "Year",
    color = "Metric"
  ) +
  facet_wrap(~Region, scales = "free_x") +
  theme_minimal()

# Display as interactive Plotly
ggplotly(chart3)
```

\

# Load and Merge Data

I integrate air pollution data with a public health dataset, both sourced from Kaggle, ensuring proper cleaning and handling of missing values.

```{r, echo=FALSE}

# Download air pollution dataset
#air_pollution <- read_csv(air_pollution_url)

# Assume health data is downloaded manually from IHME and loaded
# Placeholder for actual data processing
mydata <- file.path("C:","Users", "prabhser", "OneDrive - Technological University Dublin", "Desktop", "Computer Science", "year 4", "sem 2", "visualising data", "assignment")

# load dataframes from respective 
air_df <- file.path(mydata, "air_pollution.csv")
health_df <- file.path(mydata, "health_outcomes.csv")

# load data from dataframes
air_pollution <- read_csv(air_df)
health_outcomes <- read_csv(health_df)

# ensure common fields are of same type for join to succeed
air_pollution$Country <- as.character(air_pollution$Country)
health_outcomes$Country <- as.character(health_outcomes$Country)

#ensure all fields are in same case
air_pollution <- air_pollution %>% 
  mutate(Country = trimws(tolower(Country)))

health_outcomes <- health_outcomes %>% 
  mutate(Country = trimws(tolower(Country)))

# Merge datasets on country and year
data <- full_join(air_pollution, health_outcomes, by = c("Country"))

#sum(is.na(air_pollution$`WHO Country Name`))
#sum(is.na(health_outcomes$Country))

#setdiff(air_pollution$`WHO Country Name`, health_outcomes$Country)
#setdiff(health_outcomes$Country, air_pollution$`WHO Country Name`)

# Handle missing values
data <- data %>% drop_na()
glimpse(data)

str(data)
summary(data)
```

# Exploratory Data Analysis

Before creating explanatory visualizations, I explored the life expectancy vs pollution to see how they compare.

```{r, echo=FALSE}
ggplot(data, aes(x = `AQI Value`, y = LifeExpectancy, color = Region)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  labs(
    title = "Air Pollution (PM2.5) vs Life Expectancy",
    x = "PM2.5 Concentration (µg/m³)",
    y = "Age",
    color = "Region"
  ) +
  theme_minimal()
```

# Interactive Dashboard

Using shiny and plotly packages, I create an interactive dashboard to visualize trends by region and year.

```{r, echo=FALSE}
#| fig-height: 8
# Interactive but not Shiny-dependent
filtered_data <- combined_data %>% 
  filter(Region == "Europe")  # Default filter

plot_ly(filtered_data, x = ~Date, y = ~`AQI Value`, 
        color = ~Country, type = 'scatter', mode = 'lines') %>%
  layout(
    updatemenus = list(
      list(
        type = "dropdown",
        buttons = list(
          list(method = "restyle",
               args = list("y", list(filtered_data$`AQI Value`)),
               label = "Show AQI"),
          list(method = "restyle",
               args = list("y", list(filtered_data$LifeExpectancy)),
               label = "Show Life Expectancy")
        )
      )
    )
  )
```

\

```{r}
# Simulated dataset (replace with actual dataset)
data <- combined_data

renderPlotly({
  filtered_data <- data %>% filter(Region == input$Region)
  p <- ggplot(filtered_data, aes(x = Date, y = `AQI Value`, color = Country)) +
    geom_line() +
    labs(title = "PM2.5 Levels Over Time", x = "Date", y = "AQI Value") +
    theme_minimal()
  ggplotly(p)
})

renderDT({
  filtered_data <- data %>% filter(Region == input$Region)
  datatable(filtered_data)
})
```

# Conclusion

The data suggests a strong correlation between air pollution levels and life expectancies. The interactive dashboard allows for further regional analysis, aiding policymakers in making data-driven decisions to mitigate pollution effects.
